Secure Composition of Robust and Optimising Compilers

### 2023-06-15 popl/main.tex
add visualisation

It is expected that compilers preserve robust satisfaction.
In this paper, compilers are modelled as function from one programming language's syntax to the other.
Since this paper considers composing compilers, we'll adopt notation similar to tombstone diagrams\MK{cite} to easen the formal burden.
So, denote a compiler from $\src{L}$ to $\trg{L}$ as $\cc{\src{L}\to\trg{L}}{\src{\bullet}}$ and refer to $\src{L}$ as \emph{source} language and to $\trg{L}$ as \emph{target} language.
Composing $\cc{\src{L}\to\trg{L}}{\src{\bullet}}$ with $\cc{\trg{L}\to\obj{L}}{\trg{\bullet}}$ is denoted as $\cc{\src{L}\to\trg{L}\to\obj{L}}{\src{\bullet}}$ and means first running $\cc{\src{L}\to\trg{L}}{\src{\bullet}}$ and plugging the output into $\cc{\trg{L}\to\obj{L}}{\trg{\bullet}}$.
Contrary to tombstones, the implementation language of a compiler is disregarded.
Now, define the compiler-property robust trace-property preservation.

\begin{definition}[Robust Trace-Property Preservation]\label{def:rtp}
  A compiler $\cc{\src{L}\to\trg{L}}{\src{\bullet}}$ robustly preserves a class of properties $\class$ ($\rtp{\cc{\src{L}\to\trg{L}}{\src{\bullet}}}{\class}$) iff for any property $\pi\in\class$ and $\src{p}$ written in $\src{L}$, given $\rsat{\src{\progvar}}{\pi}$, then $\rsat{\cc{\src{L}\to\trg{L}}{\src{\progvar}}}{\pi}$.
\end{definition}

@@

\section{Composing Secure Compilers}\label{sec:compcomp}
This section studies different kinds of secure compiler compositions and their corresponding proofs.
Most notably, it looks at the sequential composition of secure compilers and shows a theorem that allows for modular proof engineering: compilers need not to robustly preserve the same property in order to be securely composed.
Moreover, it demonstrates that the order does not matter for composing compilers with the same source and target language.
Throughout this section, let $\src{L},\trg{L},$ and $\obj{L}$ be arbitrary programming lanugages.

\begin{definition}[Sequential Composition]\label{def:seqcomp}
  Let $\cc{\src{L}\to\trg{L}}{\src{\bullet}}$, $\cc{\trg{L}\to\obj{L}}{\trg{\bullet}}$ be compilers.
  Let $\src{p}$ be any $\src{L}$ program, $\trg{p}=\cc{\src{L}\to\trg{L}}{\src{p}}$, and $\obj{p}=\cc{\trg{L}\to\obj{L}}{\trg{p}}$.
  The sequential composition is $\cc{\src{L}\to\trg{L}\to\obj{L}}{\src{p}}=\obj{p}$.
\end{definition}

\begin{theorem}[Sequential Composition of Secure Compilers]\label{thm:rtp}
  Given classes of properties $\class[_{1}],\class[_{2}]$, let $\rtp{\cc{\src{L}\to\trg{L}}{\src{\bullet}}}{\class[_{1}]}$ and $\rtp{\cc{\trg{L}\to\obj{L}}{\trg{\bullet}}}{\class[_{2}]}$.
  Then, $\rtp{\cc{\src{L}\to\trg{L}\to\obj{L}}{\src{\bullet}}}{\class[_{1}]\cap\class[_{2}]}$. \Coqed
\end{theorem}

A corollary of \Cref{thm:rtp} is that optimization passes that have the input and output language can be freely swapped:
\begin{corollary}[Sequential Composition of Secure Compilers]\label{corr:swappable}
  Given classes of properties $\class[_{1}],\class[_{2}]$, let $\rtp{\cc{\trg{L}\xrightarrow{1}\trg{L}}{\trg{\bullet}}}{\class[_{1}]}$ and $\rtp{\cc{\trg{L}\xrightarrow{2}\trg{L}}{\trg{\bullet}}}{\class[_{2}]}$.
  Then, $\rtp{\cc{\trg{L}\xrightarrow{1}\trg{L}\xrightarrow{2}\trg{L}}{\trg{\bullet}}}{\class[_{1}]\cap\class[_{2}]}$ and $\rtp{\cc{\trg{L}\xrightarrow{2}\trg{L}\xrightarrow{1}\trg{L}}{\trg{\bullet}}}{\class[_{2}]\cap\class[_{1}]}$. \Coqed
\end{corollary}

\section{Case Study: Memory Safe and Cryptographic Constant Time Code}\label{sec:casestud}
This section defines several secure compilers, each of which robustly preserves a different property of interest.
The power of the frameworks from \Cref{sec:compprop,sec:compcomp} is demonstrated by composing these compilers for a secure, but optimizing, compilation chain that robustly preserves \gls{ms} and \gls{cct}.
Concretely, the section starts starts by defining languages $\src{L_{TMS}}$ and $\trg{L}$ and describes a compiler $\cc{\src{L_{TMS}}\to\trg{L}}{\src{\bullet}}$ for the robust preservation of \gls{tms} between the two languages (\Cref{subsec:cs:tms}).
Then, it defines $\irl{L_{MS}}$ and a compiler $\cc{\trg{L}\to\irl{L_{MS}}}{\trg{\bullet}}$ that robustly preserves \gls{sms} (\Cref{subsec:cs:ms}).
By the framework presented in \Cref{sec:compcomp,sec:compprop}, a compiler $\cc{\src{L_{TMS}}\to\trg{L}\to\irl{L_{MS}}}{\src{\bullet}}$ for the robust preservation of \gls{ms} by composing $\cc{\src{L_{TMS}}\to\trg{L}}{\src{\bullet}}$ and $\cc{\trg{L}\to\irl{L_{MS}}}{\trg{\bullet}}$ is attained.
Two secure compilers $\cc{{\irl{L_{MS}}\xrightarrow{\gls{dce}}\irl{L_{MS}}}}{\irl{\bullet}}$ and $\cc{\irl{L_{MS}}\xrightarrow{\gls{cf}}\irl{L_{MS}}}{\irl{\bullet}}$ are defined that perform \gls{dce} and \gls{cf}, respectively, to model an optimzing pipeline happening somewhere in the compilation chain (\Cref{subsec:cs:opts}).
Finally, the paper defines $\obj{L_{CCT}}$ and a compiler $\cc{\irl{L_{MS}}\to\obj{L_{CCT}}}{\irl{\bullet}}$ that robustly preserves \gls{cct} (\Cref{subsec:cs:scct}).
Composing it with $\cc{\src{L_{TMS}}\to\irl{L_{MS}}}{\src{\bullet}}$ and the optimizing compilers $\cc{\irl{L_{MS}}\xrightarrow{\gls{dce}}\irl{L_{MS}}}{\irl{\bullet}}$ and $\cc{\irl{L_{MS}}\xrightarrow{\gls{cf}}\irl{L_{MS}}}{\irl{\bullet}}$ yields $\cc{\src{L_{TMS}}\to\trg{L}\to\irl{L_{MS}}\xrightarrow{\gls{cf}}\irl{L_{MS}}\xrightarrow{\gls{dce}}\irl{L_{MS}}\to\obj{L_{CCT}}}{\src{\bullet}}$, a compiler that robustly preserves a combination of \gls{ms} and \gls{cct}.

>>>>

It is expected that compilers preserve robust satisfaction.
In this paper, compilers are modelled as function from one programming language's syntax to the other.
Notationally, $\cc{\src{L}}{\trg{L}}$ denotes a compiler from $\src{L}$ to $\trg{L}$ and the paper refers to $\src{L}$ as \emph{source} language and to $\trg{L}$ as \emph{target} language.
Now, define the compiler-property robust trace-property preservation.

\begin{definition}[Robust Trace-Property Preservation]\label{def:rtp}
  A compiler $\cc{\src{L}}{\trg{L}}$ robustly preserves a class of properties $\class$ ($\rtp{\cc{\src{L}}{\trg{L}}}{\class}$) iff for any property $\pi\in\class$ and $\src{p}$ written in $\src{L}$, given $\rsat{\src{\progvar}}{\pi}$, then $\rsat{\cc{\src{L}}{\trg{L}}\left(\src{p}\right)}{\pi}$.
\end{definition}

@@

\section{Composing Secure Compilers}\label{sec:compcomp}

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
      \node (S) {$\src{L_{\tmssafe}}$};
      \node[right=1.0 of S] (T) {$\trg{L}$};
      \node[right=1.0 of T] (M) {$\irl{L_{\mssafe}}$};
      \node[below right=0.5 and 1.0 of M] (D0) {$\irl{L_{\mssafe}}$};
      \node[above right=0.5 and 1.0 of M] (C0) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of D0] (D1) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of C0] (C1) {$\irl{L_{\mssafe}}$};
      \node[right=4.5 of M] (O) {$\obj{L_{\scctsafe}}$};

      \draw[->] (S) to[sloped,above] node {\text{\scriptsize\gls{tms}}} (T);
      \draw[->] (T) to[sloped,above] node {\text{\scriptsize\gls{sms}}} (M);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (D0);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (C0);
      \draw[->] (D0) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (D1);
      \draw[->] (C0) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (C1);
      \draw[->] (D1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);
      \draw[->] (C1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimizing compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}
\end{figure}

This section studies different kinds of secure compiler compositions and their corresponding proofs.
Most notably, it looks at the sequential composition of secure compilers and shows a theorem that allows for modular proof engineering: compilers need not to robustly preserve the same property in order to be securely composed.
Moreover, it demonstrates that the order does not matter for composing compilers with the same source and target language.
Throughout this section, let $\src{L},\trg{L},$ and $\obj{L}$ be arbitrary programming lanugages.

\begin{definition}[Sequential Composition]\label{def:seqcomp}
  Let $\cc{\src{L}}{\trg{L}}$, $\cc{\trg{L}}{\obj{L}}$ be compilers.
  Let $\src{p}$ be any $\src{L}$ program, $\trg{p}=\cc{\src{L}}{\trg{L}}\left(\src{p}\right)$, and $\obj{p}=\cc{\trg{L}}{\obj{L}}\left(\trg{p}\right)$.
  The sequential composition is $\left(\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}\right)\left(\src{p}\right)=\obj{p}$.
\end{definition}

\begin{theorem}[Sequential Composition of Secure Compilers]\label{thm:rtp}
  Given classes of properties $\class[_{1}],\class[_{2}]$, let $\rtp{\cc{\src{L}}{\trg{L}}}{\class[_{1}]}$ and $\rtp{\cc{\trg{L}}{\obj{L}}}{\class[_{2}]}$.
  Then, $\rtp{\cc{\src{L}}{\trg{L}}\circ\cc{\trg{L}}{\obj{L}}}{\class[_{1}]\cap\class[_{2}]}$. \Coqed
\end{theorem}

A corollary of \Cref{thm:rtp} is that optimization passes that have the input and output language can be freely swapped:
\begin{corollary}[Sequential Composition of Secure Compilers]\label{corr:swappable}
  Given classes of properties $\class[_{1}],\class[_{2}]$, let $\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{1}]}$ and $\rtp{\cc[_{2}]{\trg{L}}{\trg{L}}}{\class[_{2}]}$.
  Then, $\rtp{\cc[_{1}]{\trg{L}}{\trg{L}}}{\class[_{1}]\cap\class[_{2}]}$ and $\rtp{\cc[_{2}]{\trg{L}}{\trg{V}}}{\class[_{2}]\cap\class[_{1}]}$. \Coqed
\end{corollary}

\section{Case Study: Memory Safe and Cryptographic Constant Time Code}\label{sec:casestud}
This section defines several secure compilers, each of which robustly preserves a different property of interest.
The power of the frameworks from \Cref{sec:compprop,sec:compcomp} is demonstrated by composing these compilers for a secure, but optimizing, compilation chain that robustly preserves \gls{ms} and \gls{cct}.
Concretely, the section starts starts by defining languages $\src{L_{TMS}}$ and $\trg{L}$ and describes a compiler $\cc{\src{L_{TMS}}}{\trg{L}}$ for the robust preservation of \gls{tms} between the two languages (\Cref{subsec:cs:tms}).
Then, it defines $\irl{L_{MS}}$ and a compiler $\cc{\trg{L}}{\irl{L_{MS}}}$ that robustly preserves \gls{sms} (\Cref{subsec:cs:ms}).
By the framework presented in \Cref{sec:compcomp,sec:compprop}, a compiler $\rtp{\cc{\src{L_{TMS}}}{\trg{L}}\circ\cc{\trg{L}}{\irl{L_{MS}}}}{\mssafe}$ is attained.
Two secure compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ are defined that perform \gls{dce} and \gls{cf}, respectively, to model an optimzing pipeline happening somewhere in the compilation chain (\Cref{subsec:cs:opts}).
Finally, the paper defines $\obj{L_{CCT}}$ and a compiler $\cc{\irl{L_{MS}}}{\obj{L_{CCT}}}$ that robustly preserves \gls{cct} (\Cref{subsec:cs:scct}).
Composing it with $\cc{\src{L_{TMS}}}{\irl{L_{MS}}}$ and the optimizing compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ yields $\cc{\src{L_{TMS}}}{\obj{L_{CCT}}}$, a compiler that robustly preserves a combination of \gls{ms} and \gls{cct}.

\MKin{add image and do introduction using that}
